---
title: "Joint Survival and QOL Modeling of PARTNER Data"
author: "Jacob Spertus"
date: ""
header-includes:
  - \usepackage{amsmath}
output: 
  beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'png', echo = FALSE, warning = FALSE, message = FALSE, caption = FALSE)
options(scipen = 20)

library(survival)
library(survminer)
library(reshape2)
library(purrr)
library(tidyr)
library(dplyr)
library(mice)
library(broom)
library(rstanarm)
library(lubridate)
library(magrittr)
library(rstan)
library(ggplot2)
source("~/Dropbox/FDA/R/Data/read_partners_data.R")
setwd("~/Dropbox/FDA/R/Joint Model/")
```

```{r configure data, include = TRUE}
#function to transform QOL to be unbounded
to_norm <- function(qol){
  qol[qol >= .99] <- .99
  qol[qol <= .01] <- .01
  qnorm(qol)
}
  
#function to generate a single imputation of missing longitudinal data using PMM (this should be improved in later iterations)
impute <- function(data){
  complete(mice(data, m = 1, method = 'pmm'))
}

#these variables must be characters for proper joinings
data_list$patientdatelist$Subject %<>% as.character()
data_list$baseline$Subject %<>% as.character()
data_list$qol_wide_cat$Subject %<>% as.character()


####################### QOL data ###################
qol_long <- data_list %>%
  pluck("qol_long") %>%
  select(Subject, startdate_ITT, FU_MDY_DT, FolderName, trialarm, time, KCCQOS) %>%
  mutate(VisitNum = as.numeric(recode(FolderName, "Baseline Visit" = 0, "30 Day Follow Up" = 1, "6 Month Follow Up" = 2, "1 Year Follow Up" = 3))) %>%
  mutate(FU_MDY_DT = mdy(format(FU_MDY_DT, "%D"))) 

#there are 21 missing KCCQOS scores
#complete case analysis here
#also do not keep baseline scores (differences in treatment are reflected in intercept)
qol_data <- qol_long %>% 
  dplyr::select(Subject, time, trialarm, KCCQOS) %>%
  #filter(time != 0) %>%
  mutate(years = time / 12) %>%
  mutate(KCCQOS = KCCQOS / 100) %>% #rescaled KCCQOS to [0,1] 
  mutate(NormalKCCQOS = to_norm(KCCQOS)) %>%
  mutate(time = time  / diff(range(time))) %>%
  mutate(treat_group = ifelse(trialarm == "TEST", "Treatment", "Control")) %>%
  na.omit() %>%
  #impute() %>%
  as_tibble() %>%
  arrange(Subject, years)

############ survival data ##############
#currently, subjects with NO QOL data are not included. Is this OK?
survival_data <- data_list[c('patientdatelist','baseline','qol_wide_cat')] %>%
  reduce(left_join, by = "Subject") %>%
  dplyr::select(Subject, DaysForITTanalysis, DeathDate, TrialArm, female, racegroup, age, bmi, height, weight, nyha, EF) %>%
  mutate(gender = ifelse(female == 1, "female", "male")) %>%
  mutate(TrialArm = ifelse(TrialArm == "TEST", 1, 0)) %>% 
  mutate(YearsForITT = (DaysForITTanalysis + 1) / 365) %>%
  mutate(Censored = is.na(DeathDate)) %>%
  select(YearsForITT, Censored, Subject, TrialArm) %>%
  filter(Subject %in% qol_data$Subject) %>% #get rid of any subjects w/o QOL data 
  arrange(Subject)
  
#time is centered
data_list_stan <- list(
  N = nrow(qol_data),
  I = n_distinct(survival_data$Subject),
  KCCQ = qol_data$NormalKCCQOS,
  Timepoint = qol_data$time,
  Treatment = survival_data$TrialArm,
  Subjects = as.numeric(as.factor(qol_data$Subject)),
  Y = survival_data$YearsForITT,
  Cen = as.numeric(survival_data$Censored)
)
```

##Sections

- Background
- Notation and Model
- MCMC Fitting
- Model Checking 
- Inference


#Why Do Joint Modeling?

- Ibrahim, Chu, and Chen 2010 (Oncology):

\begin{enumerate}
\item More efficient treatment effect on survival estimates
\item More efficient effect on longitudinal QOL estimates
\item Reduced bias on "overall" treatment effect (survival and longitudinal)
\item \textcolor{red}{Incorporating survival information into longitudinal estimates accounts for informative missing data due to death (Gould et al, 2015)}
\end{enumerate}

These qualities mean that an RCT has more power, so it can rely on a smaller sample size to achieve a given level of precision.

#Past Focuses
- Much past research has focused on reducing bias and variance of the estimated effect on survival (Gould et al, 2015) ($\beta_T$)
- A few researchers have focused more on longitudinal component (McCardle et al, 2005) ($\gamma_0$ and $\gamma_1$)
- Interest has recently surrounded magnitude of association between longitudinal and survival components ($\theta_{0i}$, $\theta_{1i}$, $\beta_\theta$)
- Much development has been in oncology and AIDS research, with past joint modeling practices and guidelines reflecting idiosyncracies of those fields.
- We would like to develop methods specifically with cardiological applications in mind.



#A QOL and Survival Model

$i \in \{1,...,I\}$ indexes subjects; $j \in \{1,..,J\}$ indexes time points. $Y_{ij}$ gives the KCCQOS score at time $j$ for subject $i$; $T_i$ is treatment; $S_i$ is survival time and is possibly right censored. 

\begin{align}
  Y_{ij} &= \mu_{\theta_0} + \theta_{0i} + \mu_{\theta_1} j + \theta_{1i} j + \gamma_0 T_i + \gamma_1 T_i j + \epsilon_{ij}\\
  S_i &\sim \mbox{Weibull}\{\alpha, \exp(-(\beta_0 + \beta_T \cdot T_i + \beta_{\theta_0} \cdot \theta_{0i} + \beta_{\theta_1} \cdot \theta_{1i}))\}
\end{align}
\footnotesize{
\begin{align}
  \epsilon_{ij} &\sim \mathcal{N}(0, \sigma^2)\\
  \theta_{0i} &\sim \mathcal{N}(0, \sigma_{\theta_0}^2)\\
  \theta_{1i} &\sim \mathcal{N}(0, \sigma_{\theta_1}^2)\\
  [\mu_{\theta_0}, \mu_{\theta_1}] &\sim \mathcal{N}(0,1)\\
  [\gamma_0, \gamma_1] &\sim \mathcal{N}(0,1)\\
  [\beta_{\theta_0}, \beta_{\theta_1}] &\sim \mathcal{N}(0,1)\\
  [\beta_T, \beta_0] &\sim \mathcal{N}(0,10)\\
  [\sigma, \sigma_{\theta_0}, \sigma_{\theta_1}] &\sim t_3^+(0,1)\\
  \alpha &\sim \mathcal{N}^+(0,10)
\end{align}
}

```{r fit model, results = "hide"}
#separate_model <- stan_model("separate_model.stan")
#separate_fit <- sampling(separate_model, data = data_list_stan, chains = 4, cores = 2, iter = 1000, control = list(adapt_delta = .99, max_treedepth = 14))
#save(separate_fit, file = "separate_fit")
load("separate_fit")
#joint_model <- stan_model("joint_model_randomslope.stan")
#joint_fit <- sampling(joint_model, data = data_list_stan, chains = 4, cores = 2, iter = 1000, control = list(adapt_delta = .99, max_treedepth = 14))
#save(joint_fit, file = "re_joint_fit_randomslope")
load("re_joint_fit_randomslope")
```


#PARTNER Data

- Transcatheter aortic valve replacement (TAVR) in patients not eligible for surgical valve replacement
- $n_1 = 179$ were randomized to TAVR and $n_0 = 179$ were randomized to standard therapy
- Quality-of-life (QOL) data collected using Kansas City Cardiomyopathy Questionnaire (KCCQ) at 0, 1, 6, and 12 months
- Survival data was also recorded, and all subjects were followed for at least 1-year after implantation
- Initial analyses showed substantially improved QOL and reduction in mortality due to treatment (Reynolds et al 2011, Leon et al 2010). 
- Hazard ratio: 0.55 (95\% CI, 0.40-0.74)
- 6-month KCCQ summary difference: 21 points (95\% CI, 15-27)


```{r model diagnostics, include = FALSE}
stan_rhat(joint_fit, bins = 30)
```




```{r get model parameters}
pars <- tidyMCMC(joint_fit, conf.int = TRUE, conf.level = .95) %>% as_tibble()
fitted_subjects_frame <- pars %>%
  filter(grepl("bounded_predictor", term)) %>%
  mutate(subjects = data_list_stan$Subjects, time = data_list_stan$Timepoint, observed = pnorm(data_list_stan$KCCQ), treatment = qol_data$trialarm) 
```


#Observed Trajectories 

The trajectories observed in the data. Monthly averages are overlayed in orange.

```{r observed trajectories}
time_means <- fitted_subjects_frame %>%
  group_by(time, treatment) %>%
  summarize(mean_observed = mean(observed), se_observed = sqrt(var(estimate) / n())) %>%
  ungroup() %>%
  mutate(lower = mean_observed - 1.96*se_observed, upper = mean_observed + 1.96*se_observed)
  
plot_observed_trajectories <- ggplot(fitted_subjects_frame, aes(x = time, y = observed)) +
  geom_line(aes(group = subjects)) +
  geom_point(aes(group = subjects)) +
  geom_pointrange(data = time_means, aes(x = time, y = mean_observed, ymin = lower, ymax = upper), size = 1.1, colour = "darkorange3") +
  geom_line(data = time_means, aes(x = time, y = mean_observed), size = 1.0, colour = "darkorange3") +
  facet_grid(~ treatment) +
  labs(title = "Observed Trajectories", x = "Months") +
  ylim(0,1)
plot_observed_trajectories
```


#Linear Predictors
Predicted trajectories for individuals estimated from the model.
```{r linear predictor}
plot_linear_predictors <- ggplot(fitted_subjects_frame, aes(x = time, y = estimate, group = subjects)) +
  geom_line() +
  geom_point() +
  #ylim(0,1) +
  facet_grid(~ treatment) +
  labs(title = "Fitted Trajectories", x = "Months") +
  ylim(0,1)
plot_linear_predictors
```


#Individual Trajectories
Predicted trajectories overlayed with observed trajectories for 16 individual subjects.

```{r individual subject trajectories}
reduced_fitted_subjects_frame <- fitted_subjects_frame %>% 
  filter(subjects %in% 1:16) %>%
  gather(key = "qol_type", value = "value", estimate, observed) %>%
  unite(subject_treatment, subjects, treatment, sep = ", ") %>%
  mutate(subject_treatment = factor(subject_treatment, levels = unique(subject_treatment)))
  

plot_trajectories_subjects <- ggplot(reduced_fitted_subjects_frame, aes(x = time, y = value, color = qol_type)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ subject_treatment) +
  labs(title = "Overplotted Subject Trajectories", x = "Months") +
  theme_grey() +
  ylim(0,1)

plot_trajectories_subjects
```

#Population Level QOL Curves
The dotted lines and points with error bars are average observed KCCQOS values with 95\% credible intervals. Solid lines are model estimates with 95\% credible intervals. 

```{r population level QOL curves}
#actual curves
observed_population_curves <- qol_data %>%
  group_by(time, trialarm) %>%
  summarize(mean_KCCQOS = mean(KCCQOS, na.rm = T), se_KCCQOS = sqrt(var(KCCQOS, na.rm = T) / n())) %>%
  ungroup() %>%
  mutate(lower_KCCQOS = mean_KCCQOS - 1.96*se_KCCQOS, upper_KCCQOS = mean_KCCQOS + 1.96*se_KCCQOS) %>%
  mutate(treat_group = ifelse(trialarm ==  "TEST", "Treatment", "Control"))
#fitted curves from model
time_grid <- seq(0,1,by=.1)
qol_draws <- extract(joint_fit, pars = c("mu_theta_1","mu_theta_0", "gamma_0", "gamma_1"))
draws_mu_theta_0 <- qol_draws$mu_theta_0
draws_mu_theta_1 <- qol_draws$mu_theta_1
draws_gamma_0 <- qol_draws$gamma_0
draws_gamma_1 <- qol_draws$gamma_1
curve_control <- pnorm(as.numeric(draws_mu_theta_0) + as.matrix(draws_mu_theta_1) %*% t(time_grid)) %>% 
  melt(varnames = c("draws","time")) %>%
  group_by(time) %>%
  summarize(mean = mean(value), lower = quantile(value, .025), upper = quantile(value, .975)) %>%
  mutate(treat_group = "Control", time = time_grid)
curve_treatment <- pnorm(as.numeric(draws_mu_theta_0) + as.matrix(draws_mu_theta_1) %*% t(time_grid) + as.numeric(draws_gamma_0) + as.matrix(draws_gamma_1) %*% t(time_grid)) %>% 
  melt(varnames = c("draws","time")) %>%
  group_by(time) %>%
  summarize(mean = mean(value), lower = quantile(value, .025), upper = quantile(value, .975)) %>%
  mutate(treat_group = "Treatment", time = time_grid)
fitted_population_qol_curves <- bind_rows(curve_treatment, curve_control)
#plot population level qol curves
plot_fitted_population_qol_curves <- ggplot(fitted_population_qol_curves, aes(x = time, y = mean)) +
  geom_pointrange(data = observed_population_curves, aes(x = time, y = mean_KCCQOS, ymin = lower_KCCQOS, ymax = upper_KCCQOS, color = treat_group)) +
  geom_line(data = observed_population_curves, aes(x = time, y = mean_KCCQOS, color = treat_group), linetype = 'dashed') +
  geom_ribbon(aes(fill = treat_group, ymax = upper, ymin = lower), alpha = .2) +
  geom_line(aes(color = treat_group)) +
  ylim(0,1) +
  xlab("Time") +
  ylab("Fitted Population Level KCCQOS")
plot_fitted_population_qol_curves
```


#Survival Curves

Observed Kaplan-Meier and estimated Weibull survival curves, for each treatment group.

```{r survival curves}
#given a scale, shape, and grid of values, draw the weibull curve
compute_weibull_curve <- function(x, shape, scale){
  curve <- exp(-(x/scale)^shape)
  curve
}
survival_grid <- seq(0,2.5,by=.01)
alpha <- pars %>% filter(term == "alpha")
beta_t <- pars %>% filter(term == "beta_t")
beta_0 <- pars %>% filter(term == "beta_0")
curve_treatment <- compute_weibull_curve(x = survival_grid, shape = alpha$estimate, scale = exp(-(beta_0$estimate + beta_t$estimate) / alpha$estimate))
curve_control <- compute_weibull_curve(x = survival_grid, shape = alpha$estimate, scale = exp(-(beta_0$estimate) / alpha$estimate))
#fitted curves from model
fitted_survival_data <- data.frame(
  fitted = 100*c(curve_treatment, curve_control), 
  treat_group = c(rep("Treatment",length(survival_grid)), rep("Control", length(survival_grid))),
  time = rep(survival_grid,2)
  )
#survival data observed in sample (for Kaplan-Meier curves)
observed_survival_data <- data.frame(Y = data_list_stan$Y, Cen = 1-data_list_stan$Cen, Treatment = ifelse(data_list_stan$Treatment == 1, "Treatment", "Control"))
#km curves
surv_object <- Surv(time = data_list_stan$Y, event = 1-data_list_stan$Cen)
km_treatments <- survfit(surv_object ~ Treatment, data = observed_survival_data)
plot_km_treatments <- ggsurvplot(km_treatments, fun = 'pct', conf.int = T) +
  geom_line(data = fitted_survival_data, aes(x = time, y = fitted, group = treat_group), linetype = "dashed")
plot_km_treatments
```


<!-- #Theta Variances -->
<!-- As a final sanity check, her are the standard error of the thetas for subjects with 1, 2, 3, and 4 QOL observations. We would expect subjects with more observations to have more certain thetas (i.e. with lower standard error), and indeed we do see this. -->

```{r random effects and subject data, include = FALSE}
subject_observations <- qol_data %>%
  group_by(Subject) %>%
  summarize(observations = n(), treatment = first(trialarm)) 
theta_0 <- pars %>%
  filter(grepl("theta_0\\[", term)) %>%
  mutate(term = gsub("\\[.*", "", term)) %>%
  bind_cols(subject_observations)
theta_1 <- pars %>%
  filter(grepl("theta_1\\[", term)) %>%
  mutate(term = gsub("\\[.*", "", term)) %>%
  bind_cols(subject_observations)
thetas <- bind_rows(theta_0, theta_1)


hist_theta_se <- ggplot(thetas, aes(std.error)) +
  geom_histogram(color = "white", fill = "black", bins = 25) +
  facet_grid(observations ~ term)
hist_theta_se
```

##Inference 

#Forest Plots for Key Parameters
Treatment parameters $\gamma_0$, $\gamma_1$ and $\beta_T$ are of key interest.

```{r plot parameters}
mu_theta_0 <- pars %>% filter(term == "mu_theta_0")
mu_theta_1 <- pars %>% filter(term == "mu_theta_1")
gamma_0 <- pars %>% filter(term == "gamma_0")
gamma_1 <- pars %>% filter(term == "gamma_1")
beta_t <- pars %>% filter(term == "beta_t")
beta_theta_0 <- pars %>% filter(term == "beta_theta_0")
beta_theta_1 <- pars %>% filter(term == "beta_theta_1")
plot(joint_fit, pars = c("gamma_0","gamma_1","beta_t","beta_theta_0","beta_theta_1"))
```


#Key Parameters


The posterior mean for $\gamma_0$ is `r round(gamma_0$estimate,2)` (95\% CI: `r round(gamma_0$conf.low,2)`, `r round(gamma_0$conf.high,2)`). Under the model, subjects in the treatment group have an average 6-month QOL of `r round(pnorm(mu_theta_0$estimate + gamma_0$estimate + mu_theta_1$estimate*.5 + gamma_1$estimate*.5),2)` compared to `r round(pnorm(mu_theta_0$estimate + mu_theta_1$estimate*.5), 2)` in the control group, a difference of `r round(pnorm(mu_theta_0$estimate + gamma_0$estimate + gamma_1$estimate*.5),2) - round(pnorm(mu_theta_0$estimate + mu_theta_1$estimate*.5), 2)`. 



On the hazard ratio scale, the posterior mean for $\beta_T$ is `r round(exp(beta_t$estimate),2)` (95\% CI: `r round(exp(beta_t$conf.low),2)`, `r round(exp(beta_t$conf.high),2)`). The posterior mean for $\beta_{\theta_0}$ is `r round(exp(beta_theta_0$estimate),2)` (95\% CI: `r round(exp(beta_theta_0$conf.low),2)`, `r round(exp(beta_theta_0$conf.high),2)`), and for $\beta_{\theta_1}$ is `r round(exp(beta_theta_1$estimate),2)` (95\% CI: `r round(exp(beta_theta_1$conf.low),2)`, `r round(exp(beta_theta_1$conf.high),2)`). Ths although the estimated hazard for $\beta_{\theta_0}$ is low, it is important to note that there is considerable uncertainty in this estimate. The estimate for $\beta_{\theta_1}$ is even more uncertain, and the credible intervals are almost completely uninformative.  

#Posterior Mean Thetas
The distributions of $\theta_0$ and $\theta_1$ appear below:

```{r distribution of thetas}
thetas <- pars %>% filter(grepl("theta_0\\[", term) | grepl("theta_1\\[", term)) %>%
  separate(term, into = c("term","index"), sep = "\\[") %>%
  mutate(index = as.numeric(gsub("\\]","",index)))
theta_histogram <- ggplot(thetas, aes(estimate)) +
  geom_histogram(color = "white", fill = "black", bins = 25) +
  facet_grid(~ term, scales = "fixed")
  
theta_histogram
```

#Inferential Conclusions
A (relatively very healthy) subject with a $\theta_{0i}$ value of 1, which puts them at a mean predicted baseline QOL of `r round(pnorm(1 + mu_theta_0$estimate), 2)` in the control group or `r round(pnorm(1 + mu_theta_0$estimate + gamma_0$estimate), 2)` in the treatment group, multiplies their hazard by `r round(exp(1 * beta_theta_0$estimate), 2)` (95\% CI: `r round(exp(1 * beta_theta_0$conf.low),2)`, `r round(exp(1* beta_theta_0$conf.high),2)`). In other words, due to their high baseline QOL, these patients have under half the risk of mortality as a patient with average QOL (with considerable uncertainty however).


$\beta_T$ tells us how treatment impacts survival. The hazard ratio for treatment of `r round(exp(beta_t$estimate),2)` (95\% CrI: `r round(exp(beta_t$conf.low),2)`,`r round(exp(beta_t$conf.high),2)`) tells us that treatment basically halves the risk of mortality over time. In the original partner trial reported in NEJM, analysis indicated a hazard (at 1 year) of .55 (95\% CI: .40, .74).

#Further Questions

\begin{enumerate}
\item R packages for fitting and inference (\texttt{rstan}, \texttt{JMBayes}, etc)
\item Other specifications: Cox survival, joint trajectory, covariates
\end{enumerate}
